---
title: "Exercises_Chapt_08"
author: "Hartmut Schaefer"
date: "2024-02-12"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r include=FALSE}
library(tidyverse)
```


# 8. Linear Regression

## 8.1 Introduction

Example for regression line

```{r message=FALSE, warning=FALSE}
# Create matrix manually and convert to tibble
ma1 <- matrix(data=c(0,
                     1,
                     2,
                     3,
                     4,
                     5,
                     3.3,
                     4.3,
                     4.5,
                     9.4,
                     13.9,
                     10.5), ncol = 2)
colnames(ma1) <- c("X", "Y")
df1 <- as_tibble(ma1)
df1

# Plot as scatter plot with regression line
df1 %>% 
  ggplot(aes(x=X, y=Y))+
  geom_point(color = "blue", size=5)+
  geom_smooth(method = "lm", se=F, color = 'red')+
  xlim(0, 5)+
  ylim(0,15)+
  labs(title = "Simple regression line")

```
\

Show details of regression:

```{r}
# Show details
model <- lm(X ~ Y, data = df1)
summary(model)
```
\

Example rom video [Link](https://www.youtube.com/watch?v=sPnS1SlpJvE)\

```{r message=FALSE, warning=FALSE}
# Using Car-data-set
df_cars <- cars

# Plot data points and linear regression (LR) line

df_cars %>% 
  ggplot(aes(x=speed, y=dist))+
  geom_point(color = "blue", size = 5)+
  geom_smooth(method = "lm", se=F, color = "red")+
  xlim(0, 27)+
  ylim(-10,125)+
  labs(title = "Simple regression line")
```
\

Details of regression line model

```{r}
# Details of LR line
model1 <- lm(dist ~ speed, data = df_cars)
summary(model1)
```
\

Append additional columns to dataset (residuals)

```{r}
# Append column for residuals
df_cars$residuals <- model1$residuals
df_cars$predicted <- model1$fitted.values
head(df_cars)
```
\

Plot residuals

```{r message=FALSE, warning=FALSE}
# Plot residuals
df_cars %>% 
  ggplot(aes(x=speed, y=residuals))+
  geom_point(color = "blue", size = 3)+
  geom_smooth(method = "lm", se=F, color = "darkgrey")+
  labs(title = "Residuals")
```
\

Calculate values according to lm-model, i.e. y = a + b * x

```{r}
# calculate values from lm model for:

## create df for values to be calculated
df_values <- data.frame(speed = c(12.5, 15.5, 17))

## calculate values
df_values$dist <- predict(model1, df_values)
df_values
```

\

## 8.1.2 Using linear regression - dataset possum

### Load data

```{r load-data}
load("~/R/TestRepoStatR/data/possum.rda")
```
\

**Relationship between total length and head length**

```{r}
# Prepare data set
df_select <- possum %>% 
  select(total_l, head_l) %>% 
  arrange(total_l)
head(df_select)

```

\

**Scatterplot for Possum: total length vs head length**

```{r message=FALSE, warning=FALSE}
df_select %>% 
  ggplot(aes(x=total_l, y=head_l))+
  geom_point(color = "blue", size = 3)+
  geom_smooth(method = "lm", se=F, color = "darkgrey")+
  labs(title = "Possums: Total length vs Head length")
```
\

\

Details of regression line model

```{r}
# Details of LR line
model_possum <- lm(head_l~total_l, data = df_select)

model_possum
```
\

Predict a value

Calculate values according to lm-model, i.e. y = a + b * x

```{r}
# predict a value from lm model:

## create df for values to be calculated
df_values <- data.frame(total_l = c(80, 89))

## calculate values
df_values$head_l <- predict(model_possum, df_values)
df_values
```
\

**Residuals**

Calculate the predicted values "y_hat" (fitted values) and residuals


```{r}
# Append column for predicted values and residuals

df_select$predicted <- model_possum$fitted.values
df_select$residuals <- model_possum$residuals

head(df_select)
```
\

Plot residuals

```{r message=FALSE, warning=FALSE}
# Plot residuals
df_select %>% 
  ggplot(aes(x=total_l, y=residuals))+
  geom_point(color = "blue", size = 3)+
  geom_smooth(method = "lm", se=F, color = "darkgrey")+
  labs(title = "Possum data: Residuals")
```
\

**Correlation coefficient for observations**

"R = 1/(n-1)[(xi-xmean)(yi-ymean)]/(sx * sy)"


```{r}
# Pearson correlation coefficient R = 1/(n-1)*[(xi-xmean)(yi-ymean)]/(sx*sy)
x <- df_select$total_l
y <- df_select$head_l

r_coe <- cor(x,y, method = "pearson")
print(paste("R = ",r_coe))

```
\
\

## 8.2 Least squares regression

### Load data

```{r}
load("~/R/TestRepoStatR/data/elmhurst.rda")
```
\

**Example Emhurst College: Relationship between family income and gift aid**

```{r}
# Prepare data set
df_select_elmhurst <- elmhurst %>% 
  select(family_income, gift_aid) %>% 
  arrange(desc(family_income))
head(df_select_elmhurst)

```
\

**Scatterplot for Elmhurst College: family income vs gift aid**

```{r message=FALSE, warning=FALSE}
df_select_elmhurst %>% 
  ggplot(aes(x=family_income, y=gift_aid))+
  geom_point(color = "blue", size = 3)+
  geom_smooth(method = "lm", se=T, color = "darkgrey")+
  labs(title = "Elmhurst College: Family income vs Gift aid")
```
\


Details of regression line model

```{r}
# Details of LR line
model_elmhurst <- lm(gift_aid ~ family_income, data = df_select_elmhurst)

model_elmhurst
summary(model_elmhurst)
```
\

Calculate mean and sd for both variables

```{r}
# Pivot df to df_longer
df_select_long <- df_select_elmhurst %>% 
  pivot_longer(cols = 1:2,
               names_to = "name",
               values_to = "value")

# summarize for mean and sd
df_select_long %>% 
  group_by(name) %>% 
  summarise(mean = mean(value),
            sd = sd(value))
```
\

Calculate correlation between both variables

```{r}
# correlation R
x <- df_select_elmhurst$family_income
y <- df_select_elmhurst$gift_aid

r_coe <- cor(x,y, method = "pearson")
print(paste("R = ",r_coe))

```
\

Calculate R-squared

```{r}
# calculate R-squared
r_squared <- r_coe^2
print(paste("R-squared = ", r_squared))
```

### 8.2.8 Categorical predictors with two levels

### Load data

```{r}
load("~/R/TestRepoStatR/data/mariokart.rda")
```
\

**Relationship between Condition and Total Price**

```{r}
# Prepare data set
df_select <- mariokart %>% 
  select(cond, total_pr) %>% 
  arrange(cond, total_pr)
head(df_select)


# rename and convert to numerical
df_select$cond <- recode(df_select$cond,
                         "used" = 0,
                         "new" = 1)


# Remove extreme values (outliers)
df_select2 <- df_select %>% 
  filter(total_pr <=100)

```
\


**Scatterplot for Mario Kart: Condition vs. Total Price**

```{r message=FALSE, warning=FALSE}
df_select2 %>% 
  ggplot(aes(x=cond, y=total_pr))+
  geom_point(color = "blue", size = 5, alpha = 0.2)+
  geom_smooth(method = "lm", se=F, color = "darkgrey")+
  labs(title = "Mario Kart: Condition vs Total Price",
       subtitle = "Used (0), New(1)")
```
\

Details of regression line model

```{r}
# Details of LR line
model_mariokart <- lm(total_pr ~ cond, data = df_select2)

model_mariokart
summary(model_mariokart)
```
\
\

## 8.4 Inference for linear regression

### 8.4.1 Midterm elections and unemployment

### Load data

```{r}
load("~/R/TestRepoStatR/data/midterms_house.rda")
```
\

**Relationship between total length and head length**

Filter out election data during Great Depression (unemp > 12%)

```{r}
# Prepare data set
df_select <- midterms_house %>% 
  select(party, unemp, house_change) %>% 
  filter(unemp <= 12) %>% 
  arrange(desc(unemp))
head(df_select)
```
\

**Scatterplot for Midterms election vs. unemployment**

```{r message=FALSE, warning=FALSE}
df_select %>% 
  ggplot(aes(x=unemp, y=house_change))+
  geom_point(aes(color=party, shape=party), size=4)+
  geom_smooth(method = "lm", se=F, color = "black", size = 0.6)+
  ylim(-30, 12)+
  xlim(2.5, 12)+
  labs(title = "Change in Midterms Election vs. Unemployment Rate")
```
\

Details of regression line model

```{r}
# Details of LR line
model_midterms <- lm(house_change ~ unemp, data = df_select)

model_midterms
summary(model_midterms)

```
\

Convert List into data frame

```{r}
# Convert list into data frame

df_model_midterms <- as_tibble(summary(model_midterms)$coefficient)
df_model_midterms

```

Calculate T-value (manually)

```{r}
# Calculate t-values
t_b0 <- as.numeric((df_model_midterms[1,1] - 0)/df_model_midterms[1,2])
t_b1 <- as.numeric((df_model_midterms[2,1] - 0)/df_model_midterms[2,2])

print(paste("t-value (b0) = ", t_b0))
print(paste("t-value (b1) = ", t_b1))
```
Calculate p-value (manually)

```{r}
# calculate p-values
f <- as.numeric(model_midterms$df.residual)

p_b0 <- 2 * (1-pt(abs(t_b0), f))
p_b1 <- 2 * (1-pt(abs(t_b1), f))

print(paste("p-value (b0) = ", p_b0))
print(paste("p-value (b1) = ", p_b1))

```
\
\

### 8.4.3 Confidence interval for a coefficient

**Example Emhurst College: Relationship between family income and gift aid**\

```{r}
# Details of LR line
model_elmhurst <- lm(gift_aid ~ family_income, data = df_select_elmhurst)
summary(model_elmhurst)
```
\

Convert List into data frame

```{r}
# Convert list into data frame

df_model_elmhurst <- as_tibble(summary(model_elmhurst)$coefficient)
df_model_elmhurst

```


Calculate confidence interval of regression model coefficients, CL = 95%

```{r}
# calculate confidence interval of b0, b1, CL = 95%
confint(model_elmhurst, level = 0.95)
```
\

## Confidence and prediction intervals for linear model 

### Example 1: Penguins

[Link](https://www.youtube.com/watch?v=TBAnovN_5fg) using dataset "penguins"\

Prepare data

```{r}
library(palmerpenguins)
data(penguins)

adelie <- penguins %>% 
  filter(species == "Adelie", !is.na(bill_length_mm)) %>% 
  select(bill_depth_mm, bill_length_mm)

head(adelie)
```
\

Plot data

```{r message=FALSE, warning=FALSE}
adelie %>% 
  ggplot(aes(x = bill_depth_mm,
             y = bill_length_mm))+
  geom_point()+
  geom_smooth(method = "lm", se=T, level = 0.95, color = "darkgrey")
```
\

Build a lm - model

```{r}
model_adelie <- lm(bill_length_mm ~ bill_depth_mm, data = adelie)
summary(model_adelie)

#plot(model_adelie)
```

\

Calculate prediction interval for y_hat (bill_length_mm)

```{r warning=FALSE}
# calculate prediction interval fit, lws and upr, CI = 95%
pred_adelie <- predict(model_adelie, 
                       interval = "prediction",
                       level = 0.95)

# combine dataset and prediction interval using "cbind" (colum bind function)
adelie_new <- cbind(adelie, pred_adelie)
head(adelie_new)

```
\

Plot prediction interval

```{r message=FALSE, warning=FALSE}
# Plot prediction interval
adelie_new %>% 
  ggplot(aes(x = bill_depth_mm)) +
  geom_point(aes(y = bill_length_mm))+
  geom_smooth(aes(y = bill_length_mm), method = "lm", level = 0.95)+
  geom_line(aes(y = upr), col = "red", linetype = "dashed")+
  geom_line(aes(y = lwr), col = "red", linetype = "dashed")+
  labs(title = "Penguin Adelie bill depth vs bill length",
       subtitle = "Confidence and prediction intervals 0.95")
  
```
\
\

### Example 2: Elmhurst College**

```{r}
load("~/R/TestRepoStatR/data/elmhurst.rda")
```
\

**Example Emhurst College: Relationship between family income and gift aid**

```{r}
# Prepare data set
df_select_elmhurst <- elmhurst %>% 
  select(family_income, gift_aid) %>% 
  arrange(desc(family_income))

```
\

**Scatterplot for Elmhurst College: family income vs gift aid**

```{r message=FALSE, warning=FALSE}
df_select_elmhurst %>% 
  ggplot(aes(x=family_income, y=gift_aid))+
  geom_point(color = "darkgrey", size = 3)+
  geom_smooth(method = "lm", se=T, color = "blue")+
  labs(title = "Elmhurst College: Family income vs Gift aid")
```
\


Details of regression line model

```{r}
# Details of LR line
model_elmhurst <- lm(gift_aid ~ family_income, data = df_select_elmhurst)
```
\

Calculate prediction interval for y_hat (gift_aid)

```{r warning=FALSE}
# calculate prediction interval fit, lws and upr, CI = 95%
pred_elmhurst <- predict(model_elmhurst, 
                       interval = "prediction",
                       level = 0.95)

# combine dataset and prediction interval using "cbind" (colum bind function)
df_select_elmhurst_new <- cbind(df_select_elmhurst, pred_elmhurst)
head(df_select_elmhurst_new)

```
\

Plot prediction interval

```{r message=FALSE, warning=FALSE}
# Plot prediction interval
df_select_elmhurst_new %>% 
  ggplot(aes(x = family_income)) +
  geom_point(aes(y = gift_aid))+
  geom_smooth(aes(y = gift_aid), method = "lm", level = 0.95)+
  geom_line(aes(y = upr), col = "red", linetype = "dashed")+
  geom_line(aes(y = lwr), col = "red", linetype = "dashed")+
  labs(title = "Elmhurst College: Gift Aid vs. Family Income",
       subtitle = "Confidence and prediction intervals 0.95")
  
```
\
\

